Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-03 sections/04-application.tex
Improve Chapter on Trending, exclude notation

@@

During the COVID-19 pandemic, the need for reliable and timely nowcasts of the pandemic's development has become apparent.
In Germany, the seven-day hospitalization rate was established as a central steering measure for COVID-19 measures in November 2021, and the imposition of severe public restrictions was based on it~\parencite{RobertKochInstitute2021}.
The Robert Koch Institute (RKI) provides daily reports on the number of new hospitalizations.
However, these reports are subject to delays and revisions in two sources.
The first source is technical delays in the reporting process, for example, due to different authorities passing the data to the RKI~\parencite{RobertKochInstitute2024}.
The second, more systematic source is the structure of the seven-day hospitalization rate.
To a given date, all the cases are allocated whose first positive test result was on that date and who were hospitalized in relation to the disease in the following.
The seven-day hospitalization rate is the average count of cases per $100,000$ inhabitants on the given date and six days before.
Thus, the final seven-day hospitalization rate can only be reported with a significant delay of up to more than 70 days.
Nevertheless, as the seven-day hospitalization rate was considered a major indicator of the pandemic's development, many organizations and institutions started to issue nowcasts, including research teams and newspapers.
To collect nowcasts of the seven-day hospitalization rate by different nowcast groups, the COVID19-Nowcasting-Hub was established~\parencite{ChairOfEconometricsAndStatisticsAtKarlsruheInstituteOfTechnology2024}.
The nowcasts contain the seven-day hospitalization rate's predictive mean, median, and quantiles.

@@

The data contains eight nowcasts from scientific and public institutions, nowcast communities, and a newspaper, using different input variables, calendar data, and length of training data.
The model structures are diverse, including Bayesian models, generalized additive models, and parametric bootstrapping.
Table~\ref{tab:app-covid-models} in the appendix lists the different abbreviations and the model's overall structure.
Using the nowcasts, two ensemble methods are constructed using the ensembles' mean or median.
We denote them by ENS-MEAN and ENS-MED.
In line with the initial study design, we consider the period from November 22, 2021, to April 29, 2022, as the evaluation period.
In contrast to~\textcite{Wolffram2023}, we use the newest data available on the true values, not the data from August 8, 2022.
\unsure{Nochmal längere Evaluation?}
We do not expand on the models' performance on specific regions or age groups in Germany and the probabilistic nowcast assessment.
Figure~\ref{fig:app-covid-true-nowcast} displays the true and nowcast data for the evaluation period.
The time comprises the fourth wave's end in December 2021 and nearly the entire fifth wave of the pandemic in Germany, lasting until May 28, 2022~\parencite{Tolksdorf2022}.
Table~\ref{tab:app-covid-rmse} summarizes the point evaluation measures for the issued mean of the different models.
The models issue same-day nowcasts for nearly all 159 days of the evaluation period~\parencite[see][Tables A2, A3, and A4]{Wolffram2023}.
The best-performing models in terms of RMSE and MAE are the ILM and RKI models.
The ensemble methods perform worse than the best models regarding the mean location.
The performance of the models is diverse, with more than twice as high RMSE values for the worst models compared to the best models.
Note that the high values for the EPI model could be driven by an exceptionally far-off value at the end of the evaluation period.

@@

Figure~\ref{fig:app-covid-cond-prob-trending-ratio-7} shows the conditional trending plots and the trending ratio over the exclusion area for the lag seven days; the respective plots for the lags one day and 14 days are shown in Figure~\ref{fig:app-covid-cond-prob-trending-ratio-1-14}.
Here, only the best models in point evaluation measures, ILM, RKI, RIVM, and ENS-MED, are shown to keep the plots readable.
If RKI or ILM issues a fall in the hospitalization rate, the probability of a fall is higher than if RIVM or ENS-MED issues a fall.
The opposite is the case for a nowcasted hospitalization rate increase, and the probability margin is higher. \todo{was soll das heißen?}
Similar observations can be made for the lag of 14 days in Figure~\ref{fig:app-covid-cond-prob-14}.
For a lag of one day, the models' conditional trending ability difference is less pronounced (see Figure~\ref{fig:app-covid-cond-prob-1}).
The RKI model is still less conclusive when issuing an increase in the hospitalization rate, while RIVM is most informative in that case.
The curves cross for an issued fall, with ENS-MED being on top for issued falls above 250.

@@

For some patients, the data includes \ac{abp} and \ac{nbp} measurements.
For an overview of blood pressure measurement methods, see \textcite{Saugel2014}.
While non-invasive blood pressure measurement methods are relatively gentle, they are less accurate than invasive methods.
For critical patients, changes in blood pressure can be crucial for the treatment.
Thus, trending assessment can be performed in addition to standard accuracy analysis~\parencite[see, e.g., ]{Mostafa2020}.
Thus, we assess the trending ability of the non-invasive blood pressure measurements compared to the invasive blood pressure measurements.
Of the 64,168 numerical data records in the database, 2,662 include at least one measurement of systolic \ac{abp} and \ac{nbp} and ... include at least measurements of \ac{abp} and \ac{nbp} at the same time.
The data records generally include measurements for one patient.

@@

Again, we exclude the 10\% of absolute differences in trending assessment.
The resulting four-quadrant plots of the mean and systolic blood pressure measurements for the different lags are shown in Figure~\ref{fig:app-mimic-4q}.
The number of points in the four-quadrant plot is smaller due to the restriction to data records with measurements of mean or systolic \ac{abp} and \ac{nbp} at the same time for two consecutive times with the specified lags.
For the systolic measurements, 290, 332, and 442 points are available for the lags of one, five, and 15 minutes; for the mean measurements, 406, 430, and 542.

>>>>

In this section, we apply the trending assessment to three different applications.
In the first application in Section~\ref{sec:application-covid}, we consider the nowcasting of the seven-day hospitalization rate for COVID-19 in Germany.
Section~\ref{sec:application-eda} applies the trending assessment to forecast methods for the number of arrivals in a large emergency department.
In the last application in Section~\ref{sec:application-measurement}, we consider the trending assessment of non-invasive blood pressure measurements compared to invasive blood pressure measurements.

@@

During the COVID-19 pandemic, the need for reliable and timely nowcasts of the pandemic's development has become apparent.
In Germany, the seven-day hospitalization rate was established as a central steering measure for COVID-19 measures in November 2021, and the imposition of severe public restrictions was based on it~\parencite{RobertKochInstitute2021}.
The Robert Koch Institute (RKI) provides daily reports on the number of new hospitalizations.
However, these reports are subject to delays and revisions in two sources.
The first source is technical delays in the reporting process, for example, due to different authorities passing the data to the RKI~\parencite{RobertKochInstitute2024}.
The second, more systematic source is the structure of the seven-day hospitalization rate.
To a given date, all the cases are allocated whose first positive test result was on that date and who were hospitalized in relation to the disease in the following.
The seven-day hospitalization rate is the average count of cases per $100,000$ inhabitants on the given date and six days before.
Thus, the final seven-day hospitalization rate can only be reported with a significant delay of up to more than 70 days.
Nevertheless, as the seven-day hospitalization rate was considered a major indicator of the pandemic's development, many organizations and institutions started to issue nowcasts, including research teams and newspapers.
To collect nowcasts of the seven-day hospitalization rate by different nowcast groups, the COVID19-Nowcasting-Hub was established~\parencite{ChairOfEconometricsAndStatisticsAtKarlsruheInstituteOfTechnology2024}.
The nowcasts contain the seven-day hospitalization rate's predictive mean, median, and other quantiles.

@@

The data contains eight nowcasts from scientific and public institutions, nowcast communities, and a newspaper, using different input variables, calendar data, and length of training data.
The model structures are diverse, including Bayesian models, generalized additive models, and parametric bootstrapping.
Table~\ref{tab:app-covid-models} in the appendix lists the different abbreviations and the model's overall structure.
Using the nowcasts, two ensemble methods are constructed using the ensembles' mean or for the large .
We denote them by ENS-MEAN and ENS-MED.
In line with the initial study design, we consider the period from November 22, 2021, to April 29, 2022, as the evaluation period.
In contrast to~\textcite{Wolffram2023}, we use the newest data available on the true values, not the data from August 8, 2022.
\unsure{Nochmal längere Evaluation?}
We do not expand on the models' performance on specific regions or age groups in Germany and the probabilistic nowcast assessment.
Figure~\ref{fig:app-covid-true-nowcast} displays the true and nowcast data for the evaluation period.
The time comprises the fourth wave's end in December 2021 and nearly the entire fifth wave of the pandemic in Germany, lasting until May 28, 2022~\parencite{Tolksdorf2022}.
Table~\ref{tab:app-covid-rmse} summarizes the point evaluation measures for the issued mean of the different models.
The models issue same-day nowcasts for nearly all 159 days of the evaluation period~\parencite[see][Tables A2, A3, and A4]{Wolffram2023}.
The best-performing models in terms of RMSE and MAE are the ILM and RKI models.
The ensemble methods perform worse than the best models regarding the mean location.
The performance of the models is diverse, with more than twice as high RMSE values for the worst models compared to the best models.
Note that the high values for the EPI model could be driven by an exceptionally far-off value at the end of the evaluation period.

@@

Figure~\ref{fig:app-covid-cond-prob-trending-ratio-7} shows the conditional trending plots and the trending ratio over the exclusion area for the lag seven days; the respective plots for the lags one day and 14 days are shown in Figure~\ref{fig:app-covid-cond-prob-trending-ratio-1-14}.
Here, only the best models in point evaluation measures, ILM, RKI, RIVM, and ENS-MED, are shown to keep the plots readable.
If RKI or ILM issues a fall in the hospitalization rate, the probability of a fall is higher than if RIVM or ENS-MED issues a fall.
The opposite is the case for a nowcasted hospitalization rate increase, and the difference between the models' performance is higher.
Similar observations can be made for the lag of 14 days in Figure~\ref{fig:app-covid-cond-prob-14}.
For a lag of one day, the models' conditional trending ability difference is less pronounced (see Figure~\ref{fig:app-covid-cond-prob-1}).
The RKI model is still less conclusive when issuing an increase in the hospitalization rate, while RIVM is most informative in that case.
The curves cross for an issued fall, with ENS-MED being on top for issued falls above 250.

@@

For some patients, the data includes \ac{abp} and \ac{nbp} measurements.
For an overview of blood pressure measurement methods, see \textcite{Saugel2014}.
While non-invasive blood pressure measurement methods are relatively gentle, they are less accurate than invasive methods.
For critical patients, changes in blood pressure can be crucial for the treatment.
Thus, trending assessment can be performed in addition to standard accuracy analysis~\parencite[see, e.g., ]{Mostafa2020}.
Thus, we assess the trending ability of the non-invasive blood pressure measurements compared to the invasive blood pressure measurements.
The database contains 64,168 numerical data records.
One data record includes all numerical measurements for one patient.
In general, the measured signals vary in length, frequency and type of measurement.
Thus, only a subset of the data contains measurements of \ac{abp} and \ac{nbp} at the same time.
2,548 include at least one measurement of systolic \ac{abp} and \ac{nbp} and 1,327 include at least one measurement of systolic \ac{abp} and \ac{nbp} at the same time; for the mean \ac{abp} and \ac{nbp}, the numbers are 2,605 and 1,516, respectively.

@@

Again, we exclude the smallest 10\% of absolute differences in trending assessment.
The resulting four-quadrant plots of the mean and systolic blood pressure measurements for the different lags are shown in Figure~\ref{fig:app-mimic-4q}.
The number of points in the four-quadrant plot is smaller due to the restriction to data records with measurements of mean or systolic \ac{abp} and \ac{nbp} at the same time for two consecutive times with the specified lags.
For the systolic measurements, 290, 332, and 442 points are available for the lags of one, five, and 15 minutes; for the mean measurements, 406, 430, and 542.

### 2024-03-03 sections/50-appendix-data-generation.tex
Update bootstrapping
Improve Chapter on Trending, exclude notation

The first dataset is generated by sequentially generating $\diffx$ and $\diffy$.
First, the $\diffxt$ are sampled as a sum of a standard normal random number and a uniform random number on $(-10, 10)$:
\begin{equation*}
    \diffxt \sim N(0, 1) + U(-10, 10) \quad t = 1, \dots, T.
\end{equation*}
Subsequently, the $\diffy$ are simulated for a constant trending ratio $k$ by
\begin{equation*}
    \diffyt = \diffxt \cdot n_t * b_t,
\end{equation*}
where $n_t$ is a truncated normal distribution with mean 1 and standard deviation 0.5, truncated at 0 and a symmetric Bernoulli random variable with parameter $k$.
For a time-varying trending ratio, the parameter $k$ is modified to have a wave-shape over time, that is,
\begin{equation*}
    k_t = 0.75 + \sin(t / T^\star \cdot 2 \pi) / 4,
\end{equation*}
where $T^\star$ is $T$ divided by the number of oscillations; in this case, 4.
For the asymmetric trending ratio, $k$ is a function of $\diffxt$,
\begin{equation*}
    k(x) = 0.5 + \min \left\{ \max \left\{ \frac{x + 5}{10}, 0  \right\} , 1 \right\} / 2.
\end{equation*}

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp1}
        \caption{Constant trending ratio.}
    \end{subfigure}\hspace{0.01\textwidth}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp1_time}
        \caption{Time-varying trending ratio}
    \end{subfigure}\hspace{0.01\textwidth}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp1_asym}
        \caption{Asymmetric trending ratio}
    \end{subfigure}\hspace{0.01\textwidth}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp2}
        \caption{Second approach}
    \end{subfigure}
    \caption{Four-quadrant plots for sample realizations of the data generation schemes of Section~\ref{sec:app-trending-data-generation}. Although the first and second plot differ over time, their difference is not discernible in the plots. The third data set's asymmetry is visible in the plot but the decrese of trending ability near 0 is not visible. }
    \label{fig:appendix_dgps}
\end{figure}

>>>>

\section{Additional material on Section~\ref{sec:trending}}\label{sec:appendix-trending}

\subsection{Data generation for Section~\ref{sec:trending}}\label{subsec:app-trending-data-generation}

The first dataset is generated by sequentially generating $\diffx$ and $\diffy$.
First, the $\diffxt$ are sampled as a sum of a standard normal random number and a uniform random number on $(-10, 10)$:
\begin{equation*}
    \diffxt \sim N(0, 1) + U(-10, 10) \quad t = 1, \dots, T.
\end{equation*}
Subsequently, the $\diffy$ are simulated for a constant trending ratio $k$ by
\begin{equation*}
    \diffyt = \diffxt \cdot n_t * b_t,
\end{equation*}
where $n_t$ is a truncated normal distribution with mean 1 and standard deviation 0.5, truncated at 0 and a symmetric Bernoulli random variable with parameter $k$.
For a time-varying trending ratio, the parameter $k$ is modified to have a wave-shape over time, that is,
\begin{equation*}
    k_t = 0.75 + \sin(t / 365.25 \cdot 2 \pi) / 4,
\end{equation*}
where $T^\star$ is $T$ divided by the number of oscillations; in this case, 4.
For the asymmetric trending ratio, $k$ is a function of $\diffxt$,
\begin{equation*}
    k(x) = 0.5 + \min \left\{ \max \left\{ \frac{x + 5}{10}, 0  \right\} , 1 \right\} / 2.
\end{equation*}

@@

The four-quadrant plots for the sample realizations of the data generation schemes are shown in Figure~\ref{fig:appendix_dgps}.

\begin{figure}
    \centering
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp1}
        \caption{Constant trending ratio.}
    \end{subfigure}\hspace{0.01\textwidth}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp1_time}
        \caption{Time-varying trending ratio}
    \end{subfigure}\hspace{0.01\textwidth}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp1_asym}
        \caption{Asymmetric trending ratio}
    \end{subfigure}\hspace{0.01\textwidth}
    \begin{subfigure}{0.24\textwidth}
        \includegraphics{plots/illustrative_examples/appendix_4q_dgp2}
        \caption{Second approach}
    \end{subfigure}
    \caption{Four-quadrant plots for sample realizations of the data generation schemes of Section~\ref{subsec:app-trending-data-generation}. Although the first and second plot differ over time, their difference is not discernible in the plots. The third data set's asymmetry is visible in the plot but the decrese of trending ability near 0 is not visible. }
    \label{fig:appendix_dgps}
\end{figure}

\subsection{Simulation study on bootstrapping confidence intervals}\label{subsec:app-trending-bootstrap}
We vary the number of available time steps $T$ to be a typical time-series value, such as 30 for daily data in a month, 52 for weekly data, 168, 365, 720, and 1024.
The considered datasets are outlined in Appendix~\ref{subsec:app-trending-data-generation}, the first dataset with asymmetric dependence.
In the calculations, the \verb|scipy| package's implementation of bootstrap confidence intervals is used~\parencite{Virtanen2020}.
The prescribed confidence level is 90 \%, and the number of bootstrap samples is $10,000$.
The share of confidence intervals covering the true values per method and $T$ are shown in Table~\ref{tab:trending_bootstrap}.
The true values of the accuracy are computed based on a dataset of size $10^8$, yielding 0.7501 and 0.7700 for the two datasets.
The computation times per method and dataset are shown in Figure~\ref{fig:trending_bootstrap_time}.
For the small sample sizes up to $T = 168$, only the \ac{bca} method keeps the confidence interval size and yields slightly wider confidence intervals.
The method's results do not differ for the larger sample sizes.
The computation time for the \ac{bca} method is slightly larger than for the other methods, but all methods have a moderate computation time.

\begin{table}
    \centering
    \begin{subtable}{.48\textwidth}
        \input{plots/illustrative_examples/bootstrap_butterfly_cis}
        \caption{First dataset}
    \end{subtable}\hspace{0.02\textwidth}
    \begin{subtable}{.48\textwidth}
        \input{plots/illustrative_examples/bootstrap_normal_cis}
        \caption{Second dataset}
    \end{subtable}
    \caption{Proportion of bootstrapping confidence intervals covering the true value of trending ratio per method and sample size $T$. The average width of the confidence interval is listed in brackets.}
    \label{tab:trending_bootstrap}
\end{table}

\begin{table}
    \centering
    \begin{subfigure}{0.48\textwidth}
        \includegraphics{plots/illustrative_examples/boxplot_comp_time_butterfly}
        \caption{First dataset}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics{plots/illustrative_examples/boxplot_comp_time_normal}
        \caption{Second dataset}
    \end{subfigure}
    \caption{Boxplot of the computation time of the different bootstrapping method and data set sizes $T$. The computation time refers to bootstrapping one confidence interval based upon $10,000$ values. Each boxplot reflects $10,000$ samples. The \ac{bca} method takes slightly longer than the other two, but the difference is negligible.}
    \label{fig:trending_bootstrap_time}
\end{table}
